The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-03-01 pkg/partykit/inst/guideliketest/DAGStat2019/slides.Rnw
further revisions, changed color palettes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2480 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 10c2bbe1c2c2de545727cb9189d0869b8b8ca57f

\SweaveOpts{engine=R, eps=FALSE, keep.source=TRUE}
<<preliminaries, echo=FALSE, results=hide>>=
set.seed(7)
# setwd("~/svn/partykit/pkg/partykit/inst/guideliketest/DAGStat2019/")
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("partykit")
library("Formula")
library("parallel")
library("lattice")
library("scales")
#source("../guidelike.R")
source("../tests_guideprune.R")
#source("../ccprune.R")

@@

## HCL palette
pal <- hcl(c(140, 80, 260, 70, 10), 300, 60)
names(pal) <- c("CTree", "CTree_max", "MOB", "GUIDE", "GUIDE_scores")

## HCL palette (light)
pallight <- hcl(c(140, 80, 260, 70, 10), 300, 60, alpha = 0.25)
names(pallight) <- c("CTree", "CTree_max", "MOB", "GUIDE", "GUIDE_scores")

#produce plots
set.seed(7)
x <- runif(50,-1,1)
e1 <- rnorm(50,0,1)
e2 <- rnorm(50,0,1)
y1 <- x*6 + e1 
y2 <- x*1 + e2
y <- c(y1,y2)

jpeg("lm1.jpeg")
plot(x,y1, ylab = "", xlab = "", xaxt="n", yaxt="n", ylim = c(-7,7))
points(x, y2)
abline(lm(y~c(x,x)), col = "red", lwd = 2)
dev.off()

jpeg("lm2.jpeg")
plot(x,y1, ylab = "", xlab = "", xaxt="n", yaxt="n", ylim = c(-7,7))
abline(lm(y1~x), col = "red", lwd = 2)
dev.off()

jpeg("lm3.jpeg")
plot(x,y2, ylab = "", xlab = "", xaxt="n", yaxt="n", ylim = c(-7,7))
abline(lm(y2~x), col = "red", lwd = 2)
dev.off()

@@

\begin{frame}%[fragile]
\frametitle{Motivation}
\center
\begin{tikzpicture}
\visible<1>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 1, minimum width=90pt, minimum height = 30pt] (n0) at (4, 5) {$Y,X,Z$};
}
\visible<2>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 1, minimum width=90pt, minimum height = 30pt] (n0) at (4, 5) {$\mathcal{M}(Y,X;\hat{\beta}$)};
}
\visible<3->{
\node[inner sep=0pt] (lm1) at (4, 4.5)
    {\includegraphics[width=0.25\textwidth]{lm1.jpeg}};
}
\visible<4>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (n1) at (2, 1) {?};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (n2) at (6, 1) {?};
\draw[-, line width=0.4pt] (3.5, 3.5) -- (2.3,1.6);
\draw[-, line width=0.4pt] (4.5, 3.5) -- (5.7,1.6);
}
\visible<5-6>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (n1) at (2, 1) {$Y_1,X_1$};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (n2) at (6, 1) {$Y_2,X_2$};
\draw[-, line width=0.4pt] (3.5, 3.5) -- (2.3,1.6) node [midway, left] {\scriptsize $Z \leq \xi$};
\draw[-, line width=0.4pt] (4.5, 3.5) -- (5.7,1.6) node [midway, right] {\scriptsize $Z > \xi$};
}
\visible<6>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (n1) at (2, 1) {$\mathcal{M}(Y_1,X_1;\hat{\beta_1}$)};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (n2) at (6, 1) {$\mathcal{M}(Y_2,X_2;\hat{\beta_2}$)};
}
\visible<7->{
\node[inner sep=0pt] (lm2) at (2,0.5)
    {\includegraphics[width=0.25\textwidth]{lm2.jpeg}};
\node[inner sep=0pt] (lm3) at (6,0.5)
    {\includegraphics[width=0.25\textwidth]{lm3.jpeg}};
\draw[-, line width=0.4pt] (3.5, 3.5) -- (2.3,1.6) node [midway, left] {\scriptsize $Z \leq \xi$};
\draw[-, line width=0.4pt] (4.5, 3.5) -- (5.7,1.6) node [midway, right] {\scriptsize $Z > \xi$};
}
\end{tikzpicture}
\end{frame}

\begin{frame}%[fragile]
\frametitle{Unbiased Recursive Partitioning}
{\bf Basic tree algorithm:}
\medskip
\begin{enumerate}
\item Fit a model to the response $Y$ (and possible covariates $X$).
\item Assess association of $Y$ (or a model-based transformation)
and each possible split variable~$Z_j$ and select the split~variable~$Z_{j^\ast}$ showing the strongest association.
\item Choose the corresponding split point leading to the highest improvement of model fit and split the data.
\item Repeat steps 1--3 recursively in each of the resulting subgroups until some stopping criterion is met.
\end{enumerate}

\vspace{1.2cm}

@@

{\bf General strategy:}
\begin{enumerate}
\item Evaluate a discrepancy measure giving information on the goodness of fit of the model for each observation and each model parameter (transformation of $Y$).
\item Apply a statistical test assessing dependency of the discrepancy measure on each possible split variable $Z_j$.
\item Select the split variable $Z_j^\ast$ showing the smallest $p$-value.
\end{enumerate}

@@

{\bf Possible discrepancy measures:}
\begin{itemize}
\item Absolute deviations from an average value
\item Residuals
\item Model scores
\item \ldots
\end{itemize}
\end{frame}

\begin{frame}%[fragile]
\frametitle{Discrepancy Measures}
As an example linear models $\mathcal{LM}(Y,X;\beta_0, \beta_1)$
fitted via ordinary least squares (OLS)
are considered with the following discrepancy measures:

@@

{\bf Residuals:}
\begin{flalign*}
\hspace{0.7cm}
r(Y,X,\hat{\beta_0},\hat{\beta_1}) = Y - \hat{\beta_0} - \hat{\beta_1} \cdot X &&
\end{flalign*}

@@

{\bf Model scores:}
\begin{flalign*}
\hspace{0.7cm}
s(Y, X, \hat{\beta_0}, \hat{\beta_1}) = 
\left(\frac{\partial r^2(Y, X, \hat{\beta_0}, \hat{\beta_1})}{\partial \beta_0} \quad , \quad  
\frac{\partial r^2(Y, X, \hat{\beta_0}, \hat{\beta_1})}{\partial \beta_1}\right) 
&&
\end{flalign*}

\pause

$${\color{red}\Downarrow}$$
$$
{\color{red} 2 \cdot r(Y,X,\hat{\beta_0},\hat{\beta_1})}
$$

@@

\begin{itemize}
\item {\only<4->{\color{lightgray}} Residuals vs.\ full model scores}
\item {\only<1-3,7->{\color{lightgray}} Binarization of residuals/scores}
\item {\only<1-6>{\color{lightgray}} Categorization of possible split variables}
\end{itemize}

@@

\only<2>{
\small
$$
s(Y, X, \hat{\beta_0}, \hat{\beta_1}) = 
\begin{pmatrix} 
\partial_{\beta_0} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} \partial_{\beta_1} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})}\\
\partial_{\beta_0} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} \partial_{\beta_1} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})}\\
\vdots & {\color{lightgray} \vdots} \\
\partial_{\beta_0} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} \partial_{\beta_1} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})}
\end{pmatrix}
$$
}
\only<3>{
\small
$$
s(Y, X, \beta_0, \beta_1) = 
\begin{pmatrix} 
\partial_{\beta_0} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & \partial_{\beta_1} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})\\
\partial_{\beta_0} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & \partial_{\beta_1} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})\\
\vdots & \vdots \\
\partial_{\beta_0} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & \partial_{\beta_1} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})
\end{pmatrix}
$$
}
\only<5-6>{
\small
$$
r(Y, X, \hat{\beta_0}, \hat{\beta_1}) = 
\begin{pmatrix}
r(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})\\
r(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})\\
\vdots \\
r(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})
\end{pmatrix}
%
\visible<6>{
\quad \Rightarrow \quad
\small
\begin{pmatrix}
1\\
0\\
\vdots \\
1
\end{pmatrix}
}
$$
}
\only<7-8>{
\small
$$
Z_j = 
\begin{pmatrix}
Z_{j1}\\
Z_{j2}\\
\vdots \\
Z_{jn}\\
\end{pmatrix}
%
\visible<8>{
\quad \Rightarrow \quad
\small
\begin{pmatrix}
3\\
1\\
\vdots \\
2
\end{pmatrix}
}
$$
}
\end{frame}

@@

tp <- function (obj, digits = 3, abbreviate = FALSE, 
                fill = c("lightgray", "white"), id = TRUE, 
                just = c("center", "top"), top = 0.85, 
                align = c("center", "left", "right"), gp = NULL, FUN = NULL, 
                height = NULL, width = NULL) 
{
  nam <- names(obj)
  extract_label <- function(node) formatinfo_node(node, FUN = FUN, 
                                                  default = c("terminal", "node"))
  maxstr <- function(node) {
    lab <- extract_label(node)
    klab <- if (is.terminal(node)) 
      ""
    else unlist(lapply(kids_node(node), maxstr))
    lab <- c(lab, klab)
    lab <- try(unlist(lapply(lab, function(x) strsplit(x, 
                                                       "\n"))), silent = TRUE)
    if (inherits(lab, "try-error")) {
      paste(rep("a", 9L), collapse = "")
    }
    else {
      return(lab[which.max(nchar(lab))])
    }
  }
  nstr <- if (is.null(width)) 
    maxstr(node_party(obj))
  else paste(rep("a", width), collapse = "")
  just <- match.arg(just[1L], c("center", "centre", "top"))
  if (just == "centre") 
    just <- "center"
  align <- match.arg(align[1L], c("center", "centre", "left", 
                                  "right"))
  if (align == "centre") 
    align <- "center"
  rval <- function(node) {
    if(node$id == 2) fill <- c("white", "blue")
    if(node$id == 3 | node$id == 4) fill <- c("white", "red")
    if(node$id == 5) fill <- c("white", "darkgreen")
    fill <- rep(fill, length.out = 2)
    lab <- extract_label(node)
    if (!is.null(gp)) {
      outer_vp <- viewport(gp = gp)
      pushViewport(outer_vp)
    }
    if (is.null(height)) 
      height <- length(lab) + 1L
    node_vp <- viewport(x = unit(0.5, "npc"), y = unit(if (just == 
                                                           "top") 
      top
      else 0.5, "npc"), just = c("center", just), width = unit(1, 
                                                               "strwidth", nstr) * 1.1, height = unit(height, "lines"), 
      name = paste("node_terminal", id_node(node), sep = ""), 
      gp = if (is.null(gp)) 
        gpar()
      else gp)
    pushViewport(node_vp)
    grid.rect(gp = gpar(fill = fill[1]))
    for (i in seq_along(lab)) grid.text(x = switch(align, 
                                                   center = unit(0.5, "npc"), left = unit(1, "strwidth", 
                                                                                          "a"), right = unit(1, "npc") - unit(1, "strwidth", 
                                                                                                                              "a")), y = unit(length(lab) - i + 1, "lines"), 
                                        lab[i], just = align)
    if (id) {
      nodeIDvp <- viewport(x = unit(0.5, "npc"), y = unit(1, 
                                                          "npc"), width = max(unit(1, "lines"), unit(1.3, 
                                                                                                     "strwidth", nam[id_node(node)])), height = max(unit(1, 
                                                                                                                                                         "lines"), unit(1.3, "strheight", nam[id_node(node)])))
      pushViewport(nodeIDvp)
      grid.rect(gp = gpar(fill = fill[2], lty = "solid"))
      grid.text(nam[id_node(node)])
      popViewport()
    }
    if (is.null(gp)) 
      upViewport()
    else upViewport(2)
  }
  return(rval)
}

@@

\begin{figure}%[t]
\setkeys{Gin}{width=1\linewidth}
\minipage{0.45\textwidth}
<<dgp_stump1, fig=TRUE, echo=FALSE, width=4, height=3>>=
#par(mar=c(3,3,0,0), oma = c(3,3,0,0))
plot(p, 
     tp_args = list(FUN = identity, width = 15), 
     ip_args = list(fill = paltrees[c(2, 3)]),
     drop_terminal = TRUE, tnex = 1, edge_panel = ep, terminal_panel = tp)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_stump2, fig=TRUE, echo=FALSE, width=4, height=3>>=
par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", 
     main = expression("varying"~beta[0]),
     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
abline(coef = c(-1,1), col = "blue", lwd = 2)
abline(coef = c(1,1), col = "red", lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha("red", 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==+delta), side = 4, col = "red", las = 1, line = 0.5, at = 3.2)
mtext(expression(beta[1]== 1), side = 4, col = "red", las = 1, line = 0.5, at = 2.2)
mtext(expression(beta[0]==-delta), side = 4, col = "blue", las = 1, line = 0.5, at = 0.2)
mtext(expression(beta[1]== 1), side = 4, col = "blue", las = 1, line = 0.5, at = -0.8)
@
\endminipage

\minipage{0.45\textwidth}
<<dgp_stump3, fig=TRUE, echo=FALSE, width=4, height=3>>=
d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
               vary_beta = "beta1", beta0 = beta0, delta = delta, xi = xi, sigma = sigma)

>>>>

\SweaveOpts{engine=R, eps=FALSE, keep.source=TRUE}
<<preliminaries, echo=FALSE, results=hide>>=
set.seed(7)
# setwd("~/svn/partykit/pkg/partykit/inst/guideliketest/DAGStat2019/")
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("partykit")
library("Formula")
library("parallel")
library("lattice")
library("scales")
library("colorspace")
#source("../guidelike.R")
source("../tests_guideprune.R")
#source("../ccprune.R")

@@

## HCL palette
pal_dgp <- qualitative_hcl(3, "Set 2")

## HCL palette (light)
pal_results <- qualitative_hcl(4, "Dark 3")

#produce plots
set.seed(7)
x <- runif(70,-1,1)
e1 <- rnorm(70,0,1)
e2 <- rnorm(70,0,1)
y1 <- x*6 + e1 
y2 <- x*1 + e2
y <- c(y1,y2)

jpeg("lm1.jpeg")
par(mar=c(3.2,3.2,0,0))
plot(x,y1, ylab = "", xlab = "", xaxt="n", yaxt="n", ylim = c(-7,7))
mtext("Y",side = 2, las = 2, line = 2, cex = 2.5, col = gray.colors(1,0.3))
mtext("X",side = 1, las = 1, line = 2, cex = 2.5, col = gray.colors(1,0.3))
points(x, y2)
abline(lm(y~c(x,x)), col = "red", lwd = 3)
dev.off()

jpeg("lm2.jpeg")
par(mar=c(3.2,3.2,0,0))
plot(x,y1, ylab = "", xlab = "", xaxt="n", yaxt="n", ylim = c(-7,7))
mtext("Y",side = 2, las = 2, line = 2, cex = 2.5, col = gray.colors(1,0.3))
mtext("X",side = 1, las = 1, line = 2, cex = 2.5, col = gray.colors(1,0.3))
abline(lm(y1~x), col = "red", lwd = 3)
dev.off()

jpeg("lm3.jpeg")
par(mar=c(3.2,3.2,0,0))
plot(x,y2, ylab = "", xlab = "", xaxt="n", yaxt="n", ylim = c(-7,7))
mtext("Y",side = 2, las = 2, line = 2, cex = 2.5, col = gray.colors(1,0.3))
mtext("X",side = 1, las = 1, line = 2, cex = 2.5, col = gray.colors(1,0.3))
abline(lm(y2~x), col = "red", lwd = 3)
dev.off()

@@

\begin{frame}%[fragile]
\frametitle{Motivation}

\vspace{-0.5cm}

\center
\begin{tikzpicture}
\visible<1-3>{
\node[inner sep=0pt] (lm1) at (4, 6)
    {\includegraphics[width=0.2\textwidth]{lm1.jpeg}};
}
\visible<2>{
\node[scale=1] at (4.15, 3.7) {other covariates $Z_1, \ldots, Z_p$?};
}
\visible<3>{
\node[inner sep=0pt] (lm2) at (2,1.7)
    {\includegraphics[width=0.2\textwidth]{lm2.jpeg}};
\node[inner sep=0pt] (lm3) at (6,1.7)
    {\includegraphics[width=0.2\textwidth]{lm3.jpeg}};
\draw[-, line width=0.2pt] (3.5,4.85) -- (2.5,3.1) node [midway, left] {\scriptsize $Z_j \leq \xi$};
\draw[-, line width=0.2pt] (4.7,4.85) -- (5.7,3.1) node [midway, right] {\scriptsize $Z_j > \xi$};
}
\visible<4->{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 1, minimum width=90pt, minimum height = 30pt] (model1) at (4.15, 5.35) {$\mathcal{M}(Y,X;\hat{\beta}$)};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (model2) at (2.15, 2.55) {$\mathcal{M}(Y_1,X_1;\hat{\beta}$)};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 1, minimum width=80pt, minimum height = 30pt] (model3) at (6.15, 2.55) {$\mathcal{M}(Y_2,X_2;\hat{\beta}$)};
\draw[-, line width=0.2pt] (3.5,4.85) -- (2.5,3.1) node [midway, left] {\scriptsize $Z_j \leq \xi$};
\draw[-, line width=0.2pt] (4.7,4.85) -- (5.7,3.1) node [midway, right] {\scriptsize $Z_j > \xi$};
}
\visible<5>{
\node[scale=1] at (4.15, 1) {$\mathcal{M}$ can also be a more general model (possibly without $X$).};
}
\end{tikzpicture}
\end{frame}

\begin{frame}%[fragile]
\frametitle{Unbiased Recursive Partitioning}
{\bf Basic tree algorithm:}
\medskip
\begin{enumerate}
\item Fit a model $\mathcal{M}(Y,X;\hat{\beta})$ to the response $Y$ (and possible covariates $X$).
\item Assess association of $Y$ (or a model-based transformation)
and each possible split variable~$Z_j$ and select the split~variable~$Z_{j^\ast}$ showing the strongest association.
\item Choose the corresponding split point leading to the highest improvement of model fit and split the data.
\item Repeat steps 1--3 recursively in each of the resulting subgroups until some stopping criterion is met.
\end{enumerate}

\vspace{0.8cm}

@@

{\bf General strategy:}
\begin{enumerate}
\item Evaluate a discrepancy measure giving information on the goodness of fit of the model for each observation and each model parameter (transformation~of~$Y$).
\item Apply a statistical test assessing dependency of the discrepancy measure to each possible split variable $Z_j$.
\item Select the split variable $Z_j^\ast$ showing the smallest $p$-value.
\end{enumerate}

@@

{\bf Possible discrepancy measures:}
\begin{itemize}
\item (Ranks of) Y
\item Residuals
\item Model scores
\item Absolute deviations from an average value
\item \ldots
\end{itemize}
\end{frame}

\begin{frame}%[fragile]
\frametitle{Discrepancy Measures}
As an example linear models $\mathcal{M}(Y,X;\beta_0, \beta_1)$
fitted via ordinary least squares~(OLS)
are considered with the following discrepancy measures:

@@

{\bf Residuals:}
\begin{flalign*}
\hspace{0.7cm}
r(y,x,\hat{\beta_0},\hat{\beta_1}) = y - \hat{\beta_0} - \hat{\beta_1} \cdot x &&
\end{flalign*}

@@

{\bf Model scores} (based on least squares):
\begin{flalign*}
\hspace{0.7cm}
s(y, x, \hat{\beta_0}, \hat{\beta_1}) = 
\left(\frac{\partial r^2(y, x, \hat{\beta_0}, \hat{\beta_1})}{\partial \beta_0} \quad , \quad  
\frac{\partial r^2(y, x, \hat{\beta_0}, \hat{\beta_1})}{\partial \beta_1}\right) 
&&
\end{flalign*}

\visible<1>{
\medskip
Alternatively: Based on log-likelihood.
}
\visible<2>{
\vspace{-1cm}
$$\hspace{-0.3cm}{\color{red}\Downarrow} \hspace{3.4cm} {\color{red}\Downarrow} $$
$$
{\color{red} -2 \cdot r(y,x,\hat{\beta_0},\hat{\beta_1})}  \hspace{0.7cm} {\color{red} -2 \cdot x \cdot r(y,x,\hat{\beta_0},\hat{\beta_1})}
$$
}
\end{frame}

@@

\begin{itemize}
\item {\only<4->{\color{lightgray}} Residuals vs.\ full model scores}
\item {\only<1-3,6->{\color{lightgray}} Binarization of residuals/scores}
\item {\only<1-5>{\color{lightgray}} Categorization of possible split variables}
\end{itemize}

@@

\only<2>{
\small
$$
s(y, x, \hat{\beta_0}, \hat{\beta_1}) = 
-2 \cdot \begin{pmatrix} 
r(y_1, x_1, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} x \cdot r(y_1, x_1, \hat{\beta_0}, \hat{\beta_1})}\\
r(y_2, x_2, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} x \cdot r(y_2, x_2, \hat{\beta_0}, \hat{\beta_1})}\\
\vdots & {\color{lightgray} \vdots} \\
r(y_n, x_n, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} x \cdot r(y_n, x_n, \hat{\beta_0}, \hat{\beta_1})}
\end{pmatrix}
%\begin{pmatrix}
%\partial_{\beta_0} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} \partial_{\beta_1} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})}\\
%\partial_{\beta_0} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} \partial_{\beta_1} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})}\\
%\vdots & {\color{lightgray} \vdots} \\
%\partial_{\beta_0} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & {\color{lightgray} \partial_{\beta_1} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})}
%\end{pmatrix}
$$
}
\only<3>{
\small
$$
s(y, x, \hat{\beta_0}, \hat{\beta_1}) = 
-2 \cdot \begin{pmatrix} 
r(y_1, x_1, \hat{\beta_0}, \hat{\beta_1}) & x \cdot r(y_1, x_1, \hat{\beta_0}, \hat{\beta_1})\\
r(y_2, x_2, \hat{\beta_0}, \hat{\beta_1}) & x \cdot r(y_2, x_2, \hat{\beta_0}, \hat{\beta_1})\\
\vdots & \vdots \\
r(y_n, x_n, \hat{\beta_0}, \hat{\beta_1}) & x \cdot r(y_n, x_n, \hat{\beta_0}, \hat{\beta_1})
\end{pmatrix}
%\begin{pmatrix} 
%\partial_{\beta_0} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1}) & \partial_{\beta_1} r^2(Y_1, X_1, \hat{\beta_0}, \hat{\beta_1})\\
%\partial_{\beta_0} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1}) & \partial_{\beta_1} r^2(Y_2, X_2, \hat{\beta_0}, \hat{\beta_1})\\
%\vdots & \vdots \\
%\partial_{\beta_0} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1}) & \partial_{\beta_1} r^2(Y_n, X_n, \hat{\beta_0}, \hat{\beta_1})
%\end{pmatrix}
$$
}
\only<4-5>{
\small
$$
r(y, x, \hat{\beta_0}, \hat{\beta_1}) = 
\begin{pmatrix}
r(y_1, x_1, \hat{\beta_0}, \hat{\beta_1})\\
r(y_2, x_2, \hat{\beta_0}, \hat{\beta_1})\\
\vdots \\
r(y_n, x_n, \hat{\beta_0}, \hat{\beta_1})
\end{pmatrix}
%
\visible<5>{
\quad \Rightarrow \quad
\small
\begin{pmatrix}
\text{pos}\\
\text{non-pos}\\
\vdots \\
\text{pos}
\end{pmatrix}
}
$$
}
\only<6-7>{
\small
$$
z_j = 
\begin{pmatrix}
z_{j1}\\
z_{j2}\\
\vdots \\
z_{jn}\\
\end{pmatrix}
%
\visible<7>{
\quad \Rightarrow \quad
\small
\begin{pmatrix}
\text{Q}3\\
\text{Q}1\\
\vdots \\
\text{Q}2
\end{pmatrix}
}
$$
}
\end{frame}

\begin{frame}
\frametitle{Combinations}

The testing strategies of CTree, MOB, and GUIDE can be obtained from the general 
framework by specific combinations of the building blocks together with the 
corresponding approximation type of the underlying null distribution.

\begin{table}[t]
\begin{tabular}{ l c c c c }
\hline 
      & Scores            & Binarization  & Categorization  & Null distribution \\ 
\hline
CTree & full model scores & --          & --            & conditional\\
MOB   & full model scores & --          & --            & unconditional\\
GUIDE & residuals         & \checkmark  & \checkmark    & unconditional\\
\hline
\end{tabular}
\end{table}
\end{frame}

@@

tp <- function (obj, digits = 3, abbreviate = FALSE, 
                fill = c("lightgray", "white"), id = TRUE, 
                just = c("center", "top"), top = 0.85, 
                align = c("center", "left", "right"), gp = NULL, FUN = NULL, 
                height = NULL, width = NULL) 
{
  nam <- names(obj)
  extract_label <- function(node) formatinfo_node(node, FUN = FUN, 
                                                  default = c("terminal", "node"))
  maxstr <- function(node) {
    lab <- extract_label(node)
    klab <- if (is.terminal(node)) 
      ""
    else unlist(lapply(kids_node(node), maxstr))
    lab <- c(lab, klab)
    lab <- try(unlist(lapply(lab, function(x) strsplit(x, 
                                                       "\n"))), silent = TRUE)
    if (inherits(lab, "try-error")) {
      paste(rep("a", 9L), collapse = "")
    }
    else {
      return(lab[which.max(nchar(lab))])
    }
  }
  nstr <- if (is.null(width)) 
    maxstr(node_party(obj))
  else paste(rep("a", width), collapse = "")
  just <- match.arg(just[1L], c("center", "centre", "top"))
  if (just == "centre") 
    just <- "center"
  align <- match.arg(align[1L], c("center", "centre", "left", 
                                  "right"))
  if (align == "centre") 
    align <- "center"
  rval <- function(node) {
    if(node$id == 2) fill <- c("white", pal_dgp[1])
    if(node$id == 3 | node$id == 4) fill <- c("white", pal_dgp[2])
    if(node$id == 5) fill <- c("white", pal_dgp[3])
    fill <- rep(fill, length.out = 2)
    lab <- extract_label(node)
    if (!is.null(gp)) {
      outer_vp <- viewport(gp = gp)
      pushViewport(outer_vp)
    }
    if (is.null(height)) 
      height <- length(lab) + 1L
    node_vp <- viewport(x = unit(0.5, "npc"), y = unit(if (just == 
                                                           "top") 
      top
      else 0.5, "npc"), just = c("center", just), width = unit(1, 
                                                               "strwidth", nstr) * 1.1, height = unit(height, "lines"), 
      name = paste("node_terminal", id_node(node), sep = ""), 
      gp = if (is.null(gp)) 
        gpar()
      else gp)
    pushViewport(node_vp)
    grid.rect(gp = gpar(fill = fill[1]))
    for (i in seq_along(lab)) grid.text(x = switch(align, 
                                                   center = unit(0.5, "npc"), left = unit(1, "strwidth", 
                                                                                          "a"), right = unit(1, "npc") - unit(1, "strwidth", 
                                                                                                                              "a")), y = unit(length(lab) - i + 1, "lines"), 
                                        lab[i], just = align)
    if (id) {
      nodeIDvp <- viewport(x = unit(0.5, "npc"), y = unit(1, 
                                                          "npc"), width = max(unit(1, "lines"), unit(1.3, 
                                                                                                     "strwidth", nam[id_node(node)])), height = max(unit(1, 
                                                                                                                                                         "lines"), unit(1.3, "strheight", nam[id_node(node)])))
      pushViewport(nodeIDvp)
      grid.rect(gp = gpar(fill = fill[2], lty = "solid"))
      grid.text(nam[id_node(node)])
      popViewport()
    }
    if (is.null(gp)) 
      upViewport()
    else upViewport(2)
  }
  return(rval)
}

@@

\begin{figure}%[t]
\setkeys{Gin}{width=1\linewidth}
\minipage{0.36\textwidth}
<<dgp_stump1, fig=TRUE, echo=FALSE, width=4, height=3>>=
#par(mar=c(3,3,0,0), oma = c(3,3,0,0))
plot(p, 
     tp_args = list(FUN = identity, width = 15), 
     ip_args = list(fill = paltrees[c(2, 3)]),
     drop_terminal = TRUE, tnex = 1, edge_panel = ep, terminal_panel = tp)
@
\endminipage
\minipage{0.36\textwidth}
<<dgp_stump2, fig=TRUE, echo=FALSE, width=4, height=3>>=
par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "", xlab = "", 
     main = expression("varying"~beta[0]),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
mtext("y", side = 2, line = 3)
mtext("x", side = 1, line = 2.5)
abline(coef = c(-1,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(1,1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = 3.2)
mtext(expression(beta[1]== 1), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = 2.2)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 0.2)
mtext(expression(beta[1]== 1), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = -0.8)
@
\endminipage

\minipage{0.36\textwidth}
<<dgp_stump3, fig=TRUE, echo=FALSE, width=4, height=3>>=
d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
               vary_beta = "beta1", beta0 = beta0, delta = delta, xi = xi, sigma = sigma)
